Network based multifactorial modelling microRNA-target interaction.

### 2019-09-30 NAR_form_article.Rmd
Figure defining

## Construction of miRNA:target network
miRNA and target pairs per line should be provided as edge list to construct the network. At each line minimum required information is expression levels of miRNA and the target. If available, additional data about factors effecting binding or efficiency of miRNA can be provided as separate columns (see Figure S1 in Supplementary Tables). After construction of the network, amount of miRNA per target is calculated and kept as edge data. Simply, a target will sequester miRNA proportional to its ratio amount among other targets. If additional criteria effecting the binding of miRNA to its target is provided, distribution of miRNA will be calculated accordingly. Target can be mRNA or any other ceRNA (circRNA, ncRNA, etc.) thus, throughout the manuscript terms target, gene and ceRNA are used interchangeably.

## Triggering perturbation and subsequent calculations
Initially, the network is assumed in steady-state (Figure \ref{fig:fig1}A) condition and needs least one trigger for initiating calculations. The trigger can be a change in expression level of one or more genes (Figure \ref{fig:fig1}B). After a trigger, the network undergoes iterative cycle of calculations at each of which distribution of miRNA in local neighborhood is recalculated (Figure \ref{fig:fig1}C). Based on new miRNA distribution, expression level of each node (i.e. ceRNA) is updated. This update results in change of expression value of common target (G4) in system. In this case, the common target acts as trigger for the other group (the targets interacted with M2 miRNA) in the network. Due to common targeted elements, the change in one neighborhood spreads to other neighborhoods (Figure \ref{fig:fig1}D), consequently have potential to affect whole network due to "ripple effect" (see Section 2 in Supplementary Materials and Methods).

\begin{figure*}[ht]
\begin{center}
\includegraphics[width=15cm,height=10cm]{Fig1.jpg}
\end{center}
\caption{Schematic presentation of mechanism of network based model. 
    \textbf{A)} In steady state, miRNAs (M: triangles) repress targets (G: circles) according to proportion of target expression level. 
    \textbf{B)} Two fold increase in transcript level of Gene2 (G2) acts as a trigger (shown in red). 
    \textbf{C)} Distribution of miRNA1 (M1) changes. 
    \textbf{D)} The change at expression of common target affects changes of proportional distribution of miRNA2 (M2). Expression values are rounded to integers for simplicity. Gray values on edges indicate initial distribution, red values indicate increase and blue values indicate decrease in distribution. Shades of circles indicate different levels of increase.}
\label{fig:fig1}
\end{figure*}

@@

## Multifactorial calculations in miRNA:target network
Interactions between miRNAs and their targets can be affected from various factors. So, our model integrates multiple factors when calculating overall miRNA activity. We classified factors into two categories. Binding factors determine interaction between miRNA and target and they alter amount of miRNA sequestered to target. Efficiency factors dictate degradation efficiency of miRNA on its target and define amount of target to be degraded. In other words, binding factors exert their influence before or during binding, efficiency factors exert their influence after binding. In the literature, binding free energy [@cao_predicting_2012 ; @helwak_mapping_2013] and seed type [@werfel_preferential_2017] in miRNA:target interactions are described as factors effecting binding affinity. Also, binding region (e.g, 5'UTR, 3'UTR, CDS) was shown to drastically affect miRNA degradation efficiency [@hausser_analysis_2013; @helwak_mapping_2013]. Since binding or degradation efficiency from literature have various and broad ranges, both binding and efficiency factors are normalized to their maximum values and scaled to [0,1] interval. The normalized values of factors are used to determine binding activity and miRNA efficiency on targets (Figure \ref{fig:fig2}). Binding affinities (activity, $Eff$) of miRNAs on each individual gene are calculated as shown in equation \eqref{eq:3}; $C_m$, miRNA expression in the group; $C_{gi}$, individual gene expression; gi, individual gene (Figure \ref{fig:fig2}C).

@@

Not all miRNA:target binding events result in degradation of target. The degradation of target by bound miRNA depends on efficiency factors such as binding region. Exact repression efficiency of miRNA is calculated according to equation \eqref{eq:4} (Figure \ref{fig:fig2}D); $RE^\prime_{gi}$, normalized values of region efficiency coefficient between miRNA and gene. The cooperative repression activity of miRNAs to their common targets is figured out as shown in Figure \ref{fig:fig2}E (see Section 3 in Supplementary Materials and Methods).

@@

We have applied our model in a real dataset for which experimental measurements of various factors were available. Expression levels of miRNA and genes in tumor and normal tissue of single patient are retrieved from TCGA Research Network https://www.cancer.gov/tcga. High-throughput experimental datasets which are provided miRNA:gene target pairs with interaction factors [@helwak_mapping_2013; @moore_mirnatarget_2015] (Detailed codes are available at Section 6 in Supplementary Materials and Methods). We have combined miRNA and gene expression datasets via miRNA:target gene dataset retrieved from published datasets [@helwak_mapping_2013; @moore_mirnatarget_2015]. Detailed description of network construction and its code is available in Supplementary Materials and Methods. SERPINE2, one of the most effective nodes, has been used as trigger in ceRNA network simulation in breast cancer patient network. Relevant steps and codes are available at Supplementary Materials and Methods.

@@

\begin{figure*}[ht]
\begin{center}
\includegraphics[width=15cm,height=8cm]{Fig3.jpg}
\end{center}
\caption{Target regulations with interaction parameters. 
     \textbf{A)} In the steady-state the repression activity of miRNAs on the targets after binding and repression efficiency. 
     \textbf{B)} The changes the repression activities after increasing of G2 expression. 
     \textbf{C)} Perturbation of primary neighborhoods of M1 miRNA (M1 miRNA group). 
     \textbf{D)} Regulation of gene expression of other gene group via triggering target (common target between M1 and M2)}
\label{fig:fig3}
\end{figure*}

## ceRNA:Target networks based on interaction factors
Earlier studies reported that miRNA regulatory interactions are affected by different parameters. For example, Xu et al. have investigated the importance of seed pairing type between miRNAs and their targets and target site location by using proteomics dataset [@xu_characterization_2014]. They have proposed that the features of binding between miRNA and target can be critical for miRNA efficiency. In addition, binding energy between miRNA and targets is a significant determinant for miRNA efficiency and it has been reported that strength of miRNA:target interactions is depended on binding energy of complexes [@breda_quantifying_2015]. Similarly, another study revealed that affinity is correlated with seed pairing of miRNA:target pairs and suggested that affinity is correlated with number of canonical seed base pairing [@bosson_endogenous_2014]. Therefore, we have integrated aforementioned interaction parameters that could be useful for predicting miRNA repression activity more accurately. The sample dataset used in Figure \ref{fig:fig1} is recalculated with additional factors Table \ref{tab:one} in effect with same trigger, two fold increase in Gene2 (Figure \ref{fig:fig3}). When the factors were taken into account in the system, miRNA efficiencies varied as shown in Figure \ref{fig:fig3}A. Although the miRNA:target expression ratios in steady-state were same in comparison with the sample dataset without factors, efficiency of binding and repression have changed.  On the other hand, changes in expression levels of G4 and G1 differ although they have same initial conditions (Detailed codes are available at Sections 3-4 in Supplementary Materials and Methods). This is due to fact that G4 is targeted by two miRNAs, M1 and M2 (Figures \ref{fig:fig3}D) and changed interaction factors.

@@

It is considered that entire miRNAs in the system affect targets according to target:total target ratio Figure \ref{fig:fig1}A, when interaction factors were not taken into account. However, in presence of interaction factors in miRNA:target interaction network (shown in Table \ref{tab:one}), miRNAs are distributed according to target:total target ratio and affinity factors in first like shown in Figure \ref{fig:fig2}C. After affinity mediated proportional distribution of miRNA expression, degradation factor is considered to specify count of repressive miRNA in pairs because entire bound miRNA target pairs might not be resulted with degradation (It is figured out shown in Figure \ref{fig:fig2}D). For this reason, miRNA1 (M1) has more weak repressive effect (shown as edge variables) on target Gene2 Figure \ref{fig:fig3}A in comparison with miRNA1 (M1):Gene2 (G2) interaction in Figure \ref{fig:fig1}A. On the other hand, it is possible that taking into account of these interaction factors could cause increasing miRNA repression activity like in miRNA2:Gene4 interaction Figure \ref{fig:fig3}A. When expression of Gene2 (G2) increased, expression values of all genes also changed differentially because of contribution of efficiency factors (Figure \ref{fig:fig3} B,C,D). We have considered that in our approach energy and seed type of pairs is significant for binding and targeted region is important for repression.

@@

In regulation of miRNA target sample system (Figure \ref{fig:fig3}), miRNA (M1) repressive efficiency on the primary triggering gene (G2) is low in steady-state. So it has been observed that the regulatory activities of miRNA on the targets are weak after the increase of miRNA target G2. When model is triggered with two fold increase in expression level of common target (G4), more prominent changes were observed in gene expression levels in network compared to changes caused by G2 increase. Furthermore, change in expression level of target gene that has strong miRNA repression efficiency resulted in evident perturbation in network. On the other hand, it was observed that Gene2 was weakly affected from change of Gene4 expression because of its weak interaction factors. Since perturbation efficiency of each gene is different, we have developed a function which screens each gene in the network for their perturbation efficiencies (Detailed codes are available at Section 4 in Supplementary Materials and Methods). On the other hand, when we applied the method on the minimal dataset (i.e. Gene4 is trigger element), Gene4 has been found to be the most efficient element in terms of number of perturbed elements and miRNA2 (M2) has been found to be causing the highest mean in expression changes.

@@

We have analysed the realistic network in context of perturbation efficiencies of nodes. This process were run with 3 fold upregulation and 10 iteration in conditions where elements that have less than five percent change are not taken into account (Codes are found at Section 6.5 in Supplementary Materials and methods). With this approach, SERPINE is found as one of the most effective gene nodes in network. We have observed that the results of simulation with SERPINE2 gene, a serine protease inhibitor, one of the overexpressed genes of breast cancer [@yang2018expression; @candia2006protease] is not convenient with tumor tissue expression values, when we compared the tumor gene expression dataset retrieved from TCGA and simulation results obtained via our approach (see Section 6.6 in Supplementary Materials and methods). Simulation results were not expected to be consistent with tumor tissue expression values because large number of abnormal regulations were observed in these tissues. So, it might not said that a single gene is responsible for entire regulations in cancer cells. Besides, the all other factors such as up/down regulation of miRNAs or other genes were ignored. Therefore, it would be more useful to test our approach with an ideal dataset that includes entire informations of gene expression values before and after the regulation of a known gene at existing tissue, and miRNA expression values of this tissue at initial conditions.

>>>>

## Construction of miRNA:target network
miRNA and target pairs per line should be provided as edge list to construct the network. At each line minimum required information is expression levels of miRNA and the target. If available, additional data about factors effecting binding or efficiency of miRNA can be provided as separate columns (see Table S1 in Supplementary Tables). After construction of the network, amount of miRNA per target is calculated and kept as edge data. Simply, a target will sequester miRNA proportional to its ratio amount among other targets. If additional criteria effecting the binding of miRNA to its target is provided, distribution of miRNA will be calculated accordingly (available at Table S6 in Supplementary Tables). Target can be mRNA or any other ceRNA (circRNA, ncRNA, etc.) thus, throughout the manuscript terms target, gene and ceRNA are used interchangeably.

## Triggering perturbation and subsequent calculations
Initially, the network is assumed in steady-state (Figure \ref{fig:fig1}A and Figure S2) condition and needs least one trigger for initiating calculations. The trigger can be a change in expression level of one or more genes (Figure \ref{fig:fig1}B and Figure S3). After a trigger, the network undergoes iterative cycle of calculations at each of which distribution of miRNA in local neighborhood is recalculated (Figure \ref{fig:fig1}C and Figure S4A). Based on new miRNA distribution, expression level of each node (i.e. ceRNA) is updated. This update results in change of expression value of common target (G4) in system. In this case, the common target acts as trigger for the other group (the targets interacted with M2 miRNA) in the network. Due to common targeted elements, the change in one neighborhood spreads to other neighborhoods (Figure \ref{fig:fig1}D and Figure S4B), consequently have potential to affect whole network due to "ripple effect" (see Section 2 in Supplementary Materials and Methods).

\begin{figure*}[ht]
\begin{center}
\includegraphics[width=15cm,height=10cm]{Fig1.jpg}
\end{center}
\caption{Schematic presentation of mechanism of network based model. 
    (\textbf{A}) In steady state, miRNAs (M: triangles) repress targets (G: circles) according to proportion of target expression level. 
    (\textbf{B}) Two fold increase in transcript level of Gene2 (G2) acts as a trigger (shown in red). 
    (\textbf{C}) Distribution of miRNA1 (M1) changes. 
    (\textbf{D)}) The change at expression of common target affects changes of proportional distribution of miRNA2 (M2). Expression values are rounded to integers for simplicity. Gray values on edges indicate initial distribution, red values indicate increase and blue values indicate decrease in distribution. Shades of circles indicate different levels of increase.}
\label{fig:fig1}
\end{figure*}

@@

## Multifactorial calculations in miRNA:target network
Interactions between miRNAs and their targets can be affected from various factors. So, our model integrates multiple factors when calculating overall miRNA activity. We classified factors into two categories. Binding factors determine interaction between miRNA and target and they alter amount of miRNA sequestered to target. Efficiency factors dictate degradation efficiency of miRNA on its target and define amount of target to be degraded. In other words, binding factors exert their influence before or during binding, efficiency factors exert their influence after binding. In the literature, binding free energy [@cao_predicting_2012; @helwak_mapping_2013] and seed type [@werfel_preferential_2017] in miRNA:target interactions are described as factors effecting binding affinity. Also, binding region (e.g, 5'UTR, 3'UTR, CDS) was shown to drastically affect miRNA degradation efficiency [@hausser_analysis_2013; @helwak_mapping_2013]. Since binding or degradation efficiency from literature have various and broad ranges, both binding and efficiency factors are normalized to their maximum values and scaled to [0,1] interval. The normalized values of factors are used to determine binding activity and miRNA efficiency on targets (Figure \ref{fig:fig2}). Binding affinities (activity, $Eff$) of miRNAs on each individual gene are calculated as shown in equation \eqref{eq:3}; $C_m$, miRNA expression in the group; $C_{gi}$, individual gene expression; gi, individual gene (Figure \ref{fig:fig2}C).

@@

Not all miRNA:target binding events result in degradation of target. The degradation of target by bound miRNA depends on efficiency factors such as binding region. Exact repression efficiency of miRNA is calculated according to equation \eqref{eq:4} (Figure \ref{fig:fig2}D); $RE^\prime_{gi}$, normalized values of region efficiency coefficient between miRNA and gene. The cooperative repression activity of miRNAs to their common targets is figured out as shown in Figure \ref{fig:fig2}E.

@@

We have applied our model in a real dataset for which experimental measurements of various factors were available. Expression levels of miRNA and genes in tumor and normal tissue of single patient are retrieved from TCGA Research Network https://www.cancer.gov/tcga. High-throughput experimental datasets which are provided miRNA:gene target pairs with interaction factors [@helwak_mapping_2013; @moore_mirnatarget_2015] (Detailed codes are available at Section 6 in Supplementary Materials and Methods). We have combined miRNA and gene expression datasets via miRNA:target gene dataset retrieved from published datasets [@helwak_mapping_2013; @moore_mirnatarget_2015] (Section 6 in Supplementary Materials and Methods and Table S4 in Supplementary Tables). SERPINE2, one of the most effective nodes, has been used as trigger in ceRNA network simulation in breast cancer patient network. Relevant steps and codes are available at Supplementary Materials and Methods.

@@

\begin{figure*}[ht]
\begin{center}
\includegraphics[width=15cm,height=8cm]{Fig3.jpg}
\end{center}
\caption{Target regulations with interaction parameters. 
     (\textbf{A}) In the steady-state the repression activity of miRNAs on the targets after binding and repression efficiency. 
     (\textbf{B}) The changes the repression activities after increasing of G2 expression. 
     (\textbf{C}) Perturbation of primary neighborhoods of M1 miRNA (M1 miRNA group). 
     (\textbf{D}) Regulation of gene expression of other gene group via triggering target (common target between M1 and M2)}
\label{fig:fig3}
\end{figure*}

## ceRNA:Target networks based on interaction factors
Earlier studies reported that miRNA regulatory interactions are affected by different parameters. For example, Xu et al. have investigated the importance of seed pairing type between miRNAs and their targets and target site location by using proteomics dataset [@xu_characterization_2014]. They have proposed that the features of binding between miRNA and target can be critical for miRNA efficiency. In addition, binding energy between miRNA and targets is a significant determinant for miRNA efficiency and it has been reported that strength of miRNA:target interactions is depended on binding energy of complexes [@breda_quantifying_2015]. Similarly, another study revealed that affinity is correlated with seed pairing of miRNA:target pairs and suggested that affinity is correlated with number of canonical seed base pairing [@bosson_endogenous_2014]. Therefore, we have integrated aforementioned interaction parameters that could be useful for predicting miRNA repression activity more accurately. The sample dataset used in Figure \ref{fig:fig1} is recalculated with additional factors Table \ref{tab:one} in effect with same trigger (Detailed codes can be found at Section 3 in Supplementary Materials and Methods), two fold increase in Gene2 (Figure \ref{fig:fig3}B, and Figure S6). When the factors were taken into account in the system, miRNA efficiencies varied as shown in Figure \ref{fig:fig3}A (Figure S5). Although the miRNA:target expression ratios in steady-state were same in comparison with the sample dataset without factors, efficiency of binding and repression have changed.  On the other hand, changes in expression levels of G4 (Detailed codes are available at Section 4 in Supplementary Materials and Methods; and Figure S8) and G1 differ although they have same initial conditions. This is due to fact that G4 is targeted by two miRNAs, M1 and M2 (Figures \ref{fig:fig3}D) and changed interaction factors.

@@

It is considered that entire miRNAs in the system affect targets according to target:total target ratio Figure \ref{fig:fig1}A, when interaction factors were not taken into account. However, in presence of interaction factors in miRNA:target interaction network (shown in Table \ref{tab:one}), miRNAs are distributed according to target:total target ratio and affinity factors in first like shown in Figure \ref{fig:fig2}C. After affinity mediated proportional distribution of miRNA expression, degradation factor is considered to specify count of repressive miRNA in pairs because entire bound miRNA target pairs might not be resulted with degradation (It is figured out shown in Figure \ref{fig:fig2}D). For this reason, miRNA1 (M1) has more weak repressive effect (shown as edge variables) on target Gene2 Figure \ref{fig:fig3}A in comparison with miRNA1 (M1):Gene2 (G2) interaction in Figure \ref{fig:fig1}A. On the other hand, it is possible that taking into account of these interaction factors could cause increasing miRNA repression activity like in miRNA2:Gene4 interaction Figure \ref{fig:fig3}A. When expression of Gene2 (G2) increased, expression values of all genes also changed differentially because of contribution of efficiency factors (Figure \ref{fig:fig3} B,C,D and Figure S6-7). We have considered that in our approach energy and seed type of pairs is significant for binding and targeted region is important for repression.

@@

In regulation of miRNA target sample system (Figure \ref{fig:fig3} and Figure S5-7), miRNA (M1) repressive efficiency on the primary triggering gene (G2) is low in steady-state. So it has been observed that the regulatory activities of miRNA on the targets are weak after the increase of miRNA target G2. When model is triggered with two fold increase in expression level of common target (G4), more prominent changes were observed in gene expression levels in network compared to changes caused by G2 increase. Furthermore, change in expression level of target gene that has strong miRNA repression efficiency resulted in evident perturbation in network. On the other hand, it was observed that Gene2 was weakly affected from change of Gene4 expression because of its weak interaction factors. Since perturbation efficiency of each gene is different, we have developed a function which screens each gene in the network for their perturbation efficiencies. When we applied the method on the minimal dataset (i.e. Gene4 is trigger element), Gene4 has been found to be the most efficient element in terms of number of perturbed elements and miRNA2 (M2) has been found to be causing the highest mean in expression changes (see Section 5 in Supplementary Materials and Methods).

@@

We have analysed the realistic network in context of perturbation efficiencies of nodes. This process were run with 3 fold upregulation and 10 iteration in conditions where elements that have less than five percent change are not taken into account (Codes are found at Section 6.5 in Supplementary Materials and methods). With this approach, SERPINE2 has been found as one of the most effective gene nodes in network. We have observed that gaining steady-state in realistic network takes a lot of time because of complex interactions, as expected, and SERPINE2 exhibit highest efficiency at iteration 31 with given conditions (see Section 6.6 in Supplementary Materials and Methos and Figure S9 in Supplementary Figures). We ran the model with two fold of iteration which is observed as most effective. We have observed that the results of simulation with SERPINE2 gene, a serine protease inhibitor, one of the overexpressed genes of breast cancer [@yang2018expression; @candia2006protease] is not convenient with tumor tissue expression values, when we compared the tumor gene expression dataset retrieved from TCGA and simulation results obtained via our approach (see Section 6.6 in Supplementary Materials and methods). Simulation results were not expected to be consistent with tumor tissue expression values because large number of abnormal regulations were observed in these tissues. So, it might not said that a single gene is responsible for entire regulations in cancer cells. Besides, the all other factors such as up/down regulation of miRNAs or other genes were ignored. Therefore, it would be more useful to test our approach with an ideal dataset that includes entire informations of gene expression values before and after the regulation of a known gene at existing tissue, and miRNA expression values of this tissue at initial conditions.
